Kernel density estimate for embedding of Riemannian manifolds

### 2022-07-05 paper/paper.Rmd
Change title, replace Figure 5 with density ranks

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embe
dding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  messages = FALSE,
  warning = FALSE,
  eval = TRUE
  # include = FALSE
)
options(tinytex.verbose = TRUE)
library(tidyverse)
# library(dimRed)
# library(reticulate)
library(here)
# library(viridis)
library(hdrcde)
# library(igraph)
# library(matrixcalc)
# library(akima)
library(car)
# library(ggforce)
# library(ks)
library(patchwork)
# library(copula)
library(plotly)
library(kableExtra)
# Jmisc::sourceAll(here::here("R/sources"))
# set.seed(1234)
```

@@

```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density of ISOMAP embedding for KDE with both variable and fixed bandwidth. The four colors and shapes represents the four gaussian kernels in the 2-D meta data. Variable bandwidth shows a strong linear positive relationship.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1.png")
```

>>>>

---
title: "Kernel density estimate for embedding of Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  messages = FALSE,
  warning = FALSE,
  eval = TRUE
  # include = FALSE
)
options(tinytex.verbose = TRUE)
library(tidyverse)
# library(dimRed)
# library(reticulate)
library(here)
# library(viridis)
library(hdrcde)
# library(igraph)
# library(matrixcalc)
# library(akima)
library(car)
# library(ggforce)
# library(ks)
library(patchwork)
# library(copula)
# library(plotly)
library(kableExtra)
# Jmisc::sourceAll(here::here("R/sources"))
# set.seed(1234)
```

@@

```{r tpisomapden, fig.align = 'center', out.width = "95%", fig.cap="Scatterplot of true density and estimated density ranks of ISOMAP embedding for KDE with both variable and fixed bandwidth. The four colors and shapes represents the four gaussian kernels in the 2-D meta data. Variable bandwidth shows a strong linear positive relationship.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak_density_comparison_isomap_riem0_1_rank.png")
```
